---
title: "Reproducible Research"
author: "Macedo, Glener Diniz"
date: "Saturday, August 15, 2015"
output: html_document
---
Reproducible Research: Peer Assessment 1

```{r}
# Verifies that the .csv file is in place
  if(!file.exists("activity.csv")){
    ## The data is in a .zip file format and has to be extracted in the working directory.
     if(!file.exists("activity.zip")){
        if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")){
            temp <- tempfile()
            download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
            unzip(temp)
            unlink(temp)        
        }   
        else{ 
            unzip(zipfile="getdata-projectfiles-UCI HAR Dataset.zip")
        }
     }
     else{ 
        unzip(zipfile="activity.zip")
     }
   }
```

# Loading the dataset
Then, the data can be read using the read.csv() function in R :
```{r}
# Loading and preprocessing the data
  data <- read.csv("activity.csv")
  attach(data)
```
Here are the first ten rows :
```{r}
   head(data, n = 10L)
```

```{r how_many_missing}
missing <- is.na(data$steps)
# How many missing
table(missing)
```

What is mean total number of steps taken per day?
1. Histogram of the total number of steps taken each day

In the following part, the missing values will be ignored. We will first compute a summary of the total number of steps taken each day and then make an histogram of the results (Figura 1)
```{r}
## Computes a summary of the total number of steps taken each day
library(ggplot2)
by_day <- aggregate(steps ~ date, 
                    data = data, sum
                   )
total.steps <- tapply(data$steps, 
                      data$date, 
                      FUN=sum, 
                      na.rm=TRUE
                     )
mean(total.steps, 
     na.rm=TRUE
    )
median(total.steps, 
       na.rm=TRUE
      )
mean.spd <- mean(total.steps, 
                 na.rm=TRUE
                )
median.spd <- median(total.steps, 
                     na.rm=TRUE
                    )
ggplot(by_day, aes(steps)) + 
geom_histogram(fill = "steelblue1", 
               colour = "blue", 
               breaks = c(0, 5000, 10000, 15000, 20000, 25000)
               ) +
labs(y = expression("Frequency")) + 
labs(x = expression("Total number of steps taken each day")) +
labs(title = expression("Figura 1"))

````

What is the average daily activity pattern?

```{r}
library(ggplot2)
averages <- aggregate(x=list(steps=data$steps), 
                      by=list(interval=data$interval),
                      FUN=mean, na.rm=TRUE
                     )
ggplot(data=averages, 
       aes(x=interval, y=steps)
      ) +
geom_line(colour = "red", size = 0.75, type = "open") +
xlab("Five-minute interval") +
ylab("Average number of steps taken")

```
On average across all the days in the dataset, the five-minute interval contains the maximum number of steps?

averages[which.max(averages$steps),]
Imputing missing values

There are many days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r}
missing <- is.na(data$steps)
# How many missing
print(table(missing))
```
All of the missing values are filled in with mean value for that five-minute interval.

# Replace each missing value with the mean value of its five-minute interval
```{r}
fill.value <- function(steps, 
                       interval
                       ) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, 
                            "steps"
                           ]
                  )
    return(filled)
}
filled.data <- data
filled.data$steps <- mapply(fill.value, 
                            filled.data$steps, 
                            filled.data$interval
                           )
```

Now, using the filled data set, let's make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps.

```{r}
total.steps <- tapply(filled.data$steps, 
                      filled.data$date, 
                      FUN=sum
                     )
qplot(total.steps, 
      binwidth=1000, 
      xlab="Total number of steps taken each day",
      ylab=deparse(substitute(Count))
     ) +
geom_histogram(fill = "steelblue1", 
               colour = "darkgreen", 
               )
mean(total.steps)
median(total.steps)
```

Mean and median values are higher after imputing missing data. The reason is that in the original data, there are some days with steps values NA for any interval. The total number of steps taken in such days are set to 0s by default. However, after replacing missing steps values with the mean steps of associated interval value, these 0 values are removed from the histogram of total number of steps taken each day.

Are there differences in activity patterns between weekdays and weekends?

First, let's find the day of the week for each measurement in the dataset. In this part, we use the dataset with the filled-in values.

```{r}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)  
    if (day %in% c("Monday", 
                   "Tuesday", 
                   "Wednesday", 
                   "Thursday", 
                   "Friday" 
                  )
       )
        return("Weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("Weekend")
    else
        stop("Invalid Date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day  <- sapply(filled.data$date, 
                           FUN=weekday.or.weekend
                          )
```
Now, let's make a panel plot containing plots of average number of steps taken on weekdays and weekends.

```{r}
averages <- aggregate(steps ~ interval + 
                              day,
                      data=filled.data, 
                      mean
                     )
ggplot(averages, 
       aes(interval, 
           steps)
      ) + 
       geom_line() + 
       facet_grid(day ~ .) +
       xlab("Five-minute interval") + 
       ylab("Number of Steps")
```
